NLIZE: A Perturbation-Driven Visual Interrogation Tool for Analyzing and Interpreting Natural Language Inference Models

### 2018-06-27 paper/itemizedReviewResponse.txt

Response to the reviews

We greatly appreciate all reviewers' valuable comments. The summary of major changes is listed below. We have itemized the 44 questions/comments and answered each individually.

Summary of major changes:
- Revised section 6 to better convey the relationship among the different scenarios and provide further clarification on the result observed in 6.3.
- Revised the introduction to remove hyperbole words, clarify the NLP vs. CNN data modality comparison, and simplify the contribution list.
- Clarified the domain experts' involvement at the beginning of section 4 and link the tasks with the application scenarios.
- Revised section 5 to better explain the user interactions and add discussion on how to interpret the visual encodings.
- Reworked the video, included a new introduction section and added a voiceover.

@@

------------------------------- Reviewer 1 -------------------------------

Q1. The usage scenarios also look interesting, especially the Scenario 3. Does the result to some extent suggest that the attention mechanism is not a very efficient component in this task? I expect the authors dig it further.

A: The domain experts indeed find this observation very interesting and suggest a preliminary interpretation (which we included in the revised version). They believe the attention provides a way to compose word semantics. Individual word semantics are yielded from the encoder and input embedding, while the composed semantics participate in classification layer. From the result of Scenario 3, we can extrapolate that the ultimate decision relies on the semantics (i.e., encodings and composed encodings in classifier). The encoder and classifier layer can swiftly adjust weights for word/composed semantics to correct prediction. But the attention layer, which only works on how individual words are aligned, affects the prediction less significantly. Recent works also support this observation in word embeddings, e.g., ELMo by Salant et al., where better word embeddings can substantially benefit a model. However, it by no means implies attention is not useful, as it serves as a way to compose word semantics although it does not directly output that. We revised section 6.3 to include this comment on the interpretation of the visualization result.

### 2018-06-27 paper/application.tex
update revised version

Interestingly, all configurations except one are concentrated around the full \emph{neutral} prediction. Referring back to the pipeline visualization, we observe that the only configuration that failed to produce the correct label is the one for which we allow only updating of the attention stage of the model.
%
%In other words, the updating of the attention stage seems to have significantly less impact on the prediction result compared to the classifier or encoder stage of the model.
%Similar observations can be obtained for most examples we have experimented with.
%\shusen{
%The domain experts comment that this observation is interesting and worth further investigation from the perspective of NLP research.
%The domain experts find this observation very interesting and suggest a preliminary interpretation. They believe the attention provides a way to compose word semantics. Individual word semantics are yielded from the encoder and input embedding, while the composed semantics participate in classification layer. From this analysis, we can extrapolate that adjusting the word alignment affects only the composed semantics but has no effect on word semantics, which often plays a significant role in a good model (as supported by recent works in word embeddings, e.g., ELMo~\cite{salant2017contextualized}). Without proper semantics (in both encoder and classifier), only adjusting alignment is not sufficient to yield good model. 
%
The domain experts find this observation very interesting and suggest a preliminary interpretation. They believe the attention provides a way to compose word semantics. Individual word semantics are yielded from the encoder and input embedding, whereas the composed semantics participate in the classification layer. From this analysis, we can extrapolate that the ultimate decision relies on the semantics (i.e., encodings and composed encodings in the classifier). The encoder and classifier layer can swiftly adjust weights for word/composed semantics to correct prediction. However, the attention layer, which only works on how individual words are aligned, affects the prediction less significantly. Recent works also support this observation in word embeddings, e.g., ELMo~\cite{salant2017contextualized}, where better word embeddings can substantially benefit a model. However, word embeddings by no means implies attention is not useful, because it serves as a way to compose word semantics.
%To interpret the result, we need to carry out a case by case analysis, as the separation of the stages depends on the individual implementation.
%In our case, the result indicates that the tunable parameters of the attention stage (the transformation before the alignment operation, see details in Appendix A) has less impact on the final prediction compared to the other two stages.
%However, this result does not necessarily imply the limitation of attention mechanism as the core alignment computation does not have tunable parameters that can be modified by the proposed optimization.
%}
%The attention layer is less sensitive compared to encoding layer and classification layer.
%When examined other sentences pairs, such an observation remains.
%This gives domain expert some hint about the significance of the role of each pipeline stage play in this model.
% which based on the natural extension of the margin-infused relaxed algorithm (MIRA) to the neural network generalize the concept of maligin combined effort of the  proposed technique

>>>>

Interestingly, all configurations except one are concentrated around the full \emph{neutral} prediction. Referring back to the pipeline visualization, we observe that the only configuration that failed to produce the correct label is the one for which we allow only updating of the attention stage of the model.
%
%In other words, the updating of the attention stage seems to have significantly less impact on the prediction result compared to the classifier or encoder stage of the model.
%Similar observations can be obtained for most examples we have experimented with.
%\shusen{
%The domain experts comment that this observation is interesting and worth further investigation from the perspective of NLP research.
%The domain experts find this observation very interesting and suggest a preliminary interpretation. They believe the attention provides a way to compose word semantics. Individual word semantics are yielded from the encoder and input embedding, while the composed semantics participate in classification layer. From this analysis, we can extrapolate that adjusting the word alignment affects only the composed semantics but has no effect on word semantics, which often plays a significant role in a good model (as supported by recent works in word embeddings, e.g., ELMo~\cite{salant2017contextualized}). Without proper semantics (in both encoder and classifier), only adjusting alignment is not sufficient to yield good model. 
%
The domain experts find this observation very interesting and suggest a preliminary interpretation. They believe the attention provides a way to compose word semantics. Individual word semantics are yielded from the encoder and input embedding, whereas the composed semantics participate in the classification layer. From this analysis, we can extrapolate that the ultimate decision relies on the semantics (i.e., encodings and composed encodings in the classifier). The encoder and classifier layer can swiftly adjust weights for word/composed semantics to correct prediction. However, the attention layer, which only works on how individual words are aligned, affects the prediction less significantly. Recent word embeddings works (e.g., ELMo~\cite{salant2017contextualized}) also support such an observation that better word embeddings can substantially benefit a model. However, the result by no means implies attention is not useful, because it serves as a way to compose word semantics.
%To interpret the result, we need to carry out a case by case analysis, as the separation of the stages depends on the individual implementation.
%In our case, the result indicates that the tunable parameters of the attention stage (the transformation before the alignment operation, see details in Appendix A) has less impact on the final prediction compared to the other two stages.
%However, this result does not necessarily imply the limitation of attention mechanism as the core alignment computation does not have tunable parameters that can be modified by the proposed optimization.
%}
%The attention layer is less sensitive compared to encoding layer and classification layer.
%When examined other sentences pairs, such an observation remains.
%This gives domain expert some hint about the significance of the role of each pipeline stage play in this model.
% which based on the natural extension of the margin-infused relaxed algorithm (MIRA) to the neural network generalize the concept of maligin combined effort of the  proposed technique

### 2018-06-27 paper/attention.tex
update revised version

As illustrated in Fig.~\ref{fig:attentionVis}, we employ graph and matrix two visual encodings for visualizing attention. In the graph attention view (Fig.~\ref{fig:attentionVis}(a)), a bipartite graph encoding is adopted, in which the edge thickness corresponds to the attention value. The color of the rectangle text block encodes the sum of all edge values connected to it (darker shade of blue corresponds to higher values).
%
The graph view is good for highlighting the most dominant alignments. However, if many attention values are high, the edges may become cluttered, leading to less effective visualization. Also, if the sentence structures between premise and hypothesis are drastically different, we are likely to see the prominent edges cross each other, which can also lead to confusing and misleading visual patterns.
%
The matrix attention view (Fig.~\ref{fig:attentionVis}(b)), despite being more verbose and less efficient in highlighting the dominant alignment, does not have similar shortcomings. However, extra effort may be required for identifying the words that correspond to high-value entries in the matrix. Together, the graph and matrix views complement each other and provide the same information from different perspectives (a related matrix/graph visualization scheme is presented in \cite{MaKenyonForbes2015} for studying the brain network). 
%As illustrated by the pink arrowed lines, we can see how the same attention value is visualized in both the graph and the matrix view.
To help the user recognize the correspondence during the exploration, we enable the linkage between highlighted actions in both views (see Fig.~\ref{fig:attentionVis}(a)(b), the attention of the two ``couple'' is highlighted).

>>>>

As illustrated in Fig.~\ref{fig:attentionVis}, we employ both graph and matrix visual encodings for visualizing attention. In the graph attention view (Fig.~\ref{fig:attentionVis}(a)), a bipartite graph encoding is adopted, in which the edge thickness corresponds to the attention value. The color of the rectangle text block encodes the sum of all edge values connected to it (darker shade of blue corresponds to higher values).
%
The graph view is good for highlighting the most dominant alignments. However, if many attention values are high, the edges may become cluttered, leading to less effective visualization. Also, if the sentence structures between premise and hypothesis are drastically different, we are likely to see the prominent edges cross each other, which can also lead to confusing and misleading visual patterns.
%
The matrix attention view (Fig.~\ref{fig:attentionVis}(b)), despite being more verbose and less efficient in highlighting the dominant alignment, does not have similar shortcomings. However, extra effort may be required for identifying the words that correspond to high-value entries in the matrix. Together, the graph and matrix views complement each other and provide the same information from different perspectives (a related matrix/graph visualization scheme is explored in \cite{MaKenyonForbes2015} for studying the brain network). 
%As illustrated by the pink arrowed lines, we can see how the same attention value is visualized in both the graph and the matrix view.
To help the user recognize the correspondence during the exploration, we enable the linkage between highlighted actions in both views (see Fig.~\ref{fig:attentionVis}(a)(b), the attention of the two ``couple'' is highlighted).

### 2018-06-27 paper/prediction.tex
update revised version

\subsection{Prediction View}
\label{sec:prediction}
For a given sentence pair, the model predicts a discrete probability distribution of the three labels (neutral, contradiction, and entailment).
%
In the prediction view, as illustrated in Fig.~\ref{fig:predictionView}(a), a prediction probability is encoded as a point in the barycentric coordinate system of the triangle.
Let $\mathbf{C_1}$,$\mathbf{C_2}$,$\mathbf{C_3}$ be the three points defining a triangle, and let $p_1$, $p_2$, $p_3$ be the probabilities the label is neutral, contradiction, or entailment. The coordinate of the prediction $\mathbf{C_p}$ in the triangle is computed as $\mathbf{C_p} = p_1\mathbf{C_1}+p_2\mathbf{C_2}+p_3\mathbf{C_3}$.
In the triangle, the distinctly colored background (gray, light gray, and light green) indicates the regions that correspond to different labels. The prediction result for the original sentence pair is represented by the larger yellow circle, whereas the smaller gray circles illustrate the perturbed sentence pairs. A density contour of the prediction is computed to emphasize the highly cluttered areas and distinguish the outliers.
%\shusen{
The pattern of the perturbed pairs' prediction directly conveys the stability of the model for the given instance. 
Here, we should also take the length of the sentence into consideration as more nouns and verbs will likely lead to more varied perturbations.
%}

@@

To perturb the prediction (\textbf{T3}) (optimization for solving the prediction perturbation is discussed in Section~\ref{sec:pipeline}), we need a way to communicate the reassignment of the predicted label. As illustrated in Fig.~\ref{fig:predictionView}(b), we integrate such an operation in the prediction view. When pressing and dragging the prediction (represented as a circle), the user is presented with the three options (E, N, C) corresponding to the labels. When the user hovers on one of the options, a dotted line is shown to indicate the newly assigned label. The reassignment is applied when the user releases the mouse while hovering on the label of choice.

>>>>

\subsection{Prediction View}
\label{sec:prediction}
For a given sentence pair, the model predicts a discrete probability distribution of the three labels (neutral, contradiction, and entailment).
%
In the prediction view, as illustrated in Fig.~\ref{fig:predictionView}(a), a prediction probability is encoded as a point in the barycentric coordinate system of the triangle.
Let $\mathbf{C_1}$,$\mathbf{C_2}$,$\mathbf{C_3}$ be the three points defining a triangle, and let $p_1$, $p_2$, $p_3$ be the probabilities the label is neutral, contradiction, or entailment. The coordinate of the prediction $\mathbf{C_p}$ in the triangle is computed as $\mathbf{C_p} = p_1\mathbf{C_1}+p_2\mathbf{C_2}+p_3\mathbf{C_3}$.
In the triangle, the distinctly colored background (gray, light gray, and light green) indicates the regions that correspond to different labels. The prediction result for the original sentence pair is represented by the larger yellow circle, whereas the smaller gray circles illustrate the perturbed sentence pairs. 
%A density contour of the prediction is computed to emphasize the highly cluttered areas and distinguish the outliers.
A density contour of the prediction is computed via kernel density estimation to emphasize the highly cluttered areas and distinguish the outliers.
%\shusen{
The pattern of the perturbed pairs' prediction directly conveys the stability of the model for the given sentence pair. 
Here, we should also take the length of the sentence into consideration as greater numbers of nouns and verbs will likely lead to more varied perturbations.
%}

@@

To perturb the prediction (\textbf{T3}) (the optimization for solving the prediction perturbation is discussed in Section~\ref{sec:pipeline}), we need a way to communicate the reassignment of the predicted label. As illustrated in Fig.~\ref{fig:predictionView}(b), we integrate such an operation in the prediction view. When pressing and dragging the prediction (represented as a circle), the user is presented with the three options (E, N, C) corresponding to the labels. When the user hovers on one of the options, a dotted line is shown to indicate the newly assigned label. The reassignment is applied when the user releases the mouse while hovering on the label of choice.
